Lattice Boltzmann study of mass transfer for complicated geometries

### 2012-01-26 Text/introduction.tex
Started working on the manuscript.

\newcommand{\beq}{\begin{equation}}
\newcommand{\feq}{\end{equation}}
\newcommand{\beqstar}{\begin{equation*}}
\newcommand{\feqstar}{\end{equation*}}                                     
\newcommand{\beqal}{\begin{equation}\begin{aligned}}
\newcommand{\feqal}{\end{aligned}\end{equation}}
\newcommand{\beqalstar}{\begin{equation*}\begin{aligned}}
\newcommand{\feqalstar}{\end{aligned}\end{equation*}}
\newcommand{\vol}{k_L\,a}
\newcommand{\lbubble}{L_{\mathrm{bubble}}}
\newcommand{\lunit}{L_{\mathrm{unit}}}
\newcommand{\lslug}{L_{\mathrm{slug}}}
\newcommand{\ububble}{U_{\mathrm{bubble}}}
\newcommand{\uliq}{U_{\mathrm{liq}}}
\newcommand{\ugas}{U_{\mathrm{gas}}}
\newcommand{\cbubble}{C_{\mathrm{bubble}}}

@@

\begin{document}
\maketitle
\begin{abstract}
This work gives the thorough procedure of how to determine the volumetric mass transfer coefficient
in the context of lattice Boltzmann simulations. The novelty of this work lies in the establishing
a thorough procedure of how to measure the volumetric mass transfer coefficient in case of mass
sources to be comparable of the representative unit cell, which accounts for the inhomogenous
situation. By examining $5$ representative cases taken from microfluidics with the capillary number
from $Ca=0.1$ to $Ca=1$ the following relationship was established {\color{red} Give main findings
here}. The work is of use to people performing mass transfer numerical simulations.
\end{abstract}

\section{Introduction}
The volumetric mass transfer coefficient is a very important characteristic for microchannels 
design \cite{kreutzer-overview}. It allows to
properly manufacture a microchannel with necessary properties to ensure that chemical
reactions are performed in the best manner.

The volumetric mass transfer coefficient correlations are represented as a function of the diffusion
coefficient, slug and bubble lengths, and bubble velocity. For example, \citet{yue-mass} established
experimental correlation for periodic bubble train as: 
\begin{equation}
\vol =\frac{2}{d_h} \Bigl(\frac{D
\ububble}{\lbubble+\lslug}\Bigr)^{0.5}
\Bigl(\frac{\lbubble}{\lbubble+\lslug}\Bigr)^{0.3},
\end{equation}
where $\vol$ is the volumetric mass transfer coefficient, $d_h$ is the hydrodynamic capillary
diameter, $\lbubble$ is the bubble length, $\lslug$ is the slug distance (between bubbles),
$\ububble$ is the bubble velocity. This correlation is within $10\%$ accuracy for experimental
results for the following range of
parameters: bubble velocity $0.4\,\mathrm{m/s}<\ububble<2\,\mathrm{m/s}$, bubble length
$1.4<\frac{\lbubble}{d_h}<6.3$, slug length $1<\frac{\lslug}{d_h}<3.2$,
hydrodynamic diameter $d_h = 400 \mathrm{\mu m}$.

Other correlations can be found in works of \citet{kreutzer-overview}
and \citet{bercic-mass}. Among a variety of experimental works extensively covered
\cite{yue-mass}, the numerical validations to the best author knowledge are sparse. One of the
reasons is that the experimental correlations are established for continous system, i.e. one does
not distinguish separate bubbles, but looks to the system as a whole measuring mass concentration
in different places. There are some works as how to avoid this approach and perform numerical
simulations for one unit cell. For example, \citet{vanbaten-circular} perform numerical simulations
for one unit cell using periodic boundary conditions and measure the volumetric mass transfer
coefficient as:
\begin{equation}
\label{main:simulation:equation}
k_L a=\frac{\text{Flux}}{\cbubble-C_{outlet}} \frac{\text{bubble surface area}}{\text{unit cell
volume}},
\end{equation}
where $C_{outlet}=\int{C U_{outlet}\mathrm{d}A}/\int{U_{outlet}\mathrm{d}A}$. 
The concentration flux is calculated as the difference between overall
average concentration taken in the whole domain ($C_{overall}=\int_{V} C \mathrm{d}V /V$)
at time
$t_1$ and time $t_2$ divided on the time difference $t_2-t_1$. The agreement between numerical
simulations and correlations of \citet{bercic-mass} was good. However, the numerical model only
accounts for circular capillaries and short contact times (the mass t), which indicate that the
bubble film is not saturated with the mass concentration. As well, authors took the approximation of
bubble shapes for small capillary numbers $Ca<0.01$ and then derived the hydrodynamics fields by
imposing a free-slip condition on the bubble surface.

This work aims at the establishing numerical procedures as to how properly obtain the volumetric
mass transfer coefficient in the interesting range of capillary numbers, i.e. $0.1\leq Ca\leq1.0$.
The comparison of volumetric mass transfer coefficient correlations with experimental correlations
and establishing the Sherwood number as a function of other non-dimensional parameters is the goal
of the future work.

We chose this type of the problem due to different reasons:
\begin{description}
\item[I] The gas holdup, i.e. the ratio of volume occupied by bubble to unit cell volume, is
significant. Therefore, one cannot assume that the mass source is a point source or one can have
homogenous picture. Therefore, to establish the volumetric mass transfer coefficient one needs to
mimic the continuos picture. One of the goals of this work is to establish certain criteria as how
to simulate it numerically.
\item[II] For small capillary numbers as $Ca<0.7$ \cite{giavedoni-numerical} there is a vortex in
the reference frame moving with
the bubble, which  influences and improves the corresponding mass transfer.  For larger capillary
numbers there is no vortex and stremalines go around the bubble. Thus, the problem is interesting
because of really different hydrodynamic patterns depending on capillary number. Note that the flow
patterns strongly depend on the number of parameters as the slug length, the thickness of the
thickness and the existence of the vortex in front of the bubble. There are some available
correlations in literature \cite{bercic-mass,kreutzer-overview} and one of the particular goal of
this work is to study mass transfer dependence on the number of parameters. 
\end{description}

The approach we want to use is the lattice Boltzmann method, a relatively new CFD
competitor emerged during last $20$ years \cite{frisch,mcnamara,HJ,HSB}. During years the method
was applied to simulate not only hydrodynamic problems \cite{yu}, but as well multiphase flows
\cite{Shan-chen:extended,swift,gunstensen}, heat transfer
\cite{yuan-thermal,zhang-thermal}, ferrofluids \cite{dellar-ferro,kuzmin-aniso}.

The mass transfer problems in the lattice Boltzmann framework were mainly addressed in series of
works of Dr. Ginzburg and co-authors
\cite{ginzburg-main,ginzburg-boundary-conditions,ginzburg-saturated-flow}. However, all these works
are of general nature to simulate the advection-diffusion equation via the lattice Boltzmann
framework. The manuscripts do not establish of how to obtain the volumetric mass transfer
coefficients. The work of \citet{inamuro-scalar-boundary} simulates heat and mass transfers in the
porous media. The work of \citet{jos-mass} simulates lateral mixing in cross-channel flow. While
two last works are focused at the mass transfer problems, both problems are of homogenuous nature
and do not guide as to how to simulate the non-homogenous problems.

The overall goal of this paper is to address the questions arised and to establish the numerical
procedure suitable to measure the volumetric mass transfer coefficient. We will utilize the results
of numerical simulations of flow
obtained
before \cite{kuzmin-binary2d}.

The paper is organized as follows. The mass transfer coefficients and volumetric mass transfer
coefficients are covered. The limiting cases from the continuos representation and their
application to the present simulation are given. The mass transfer simulations benchmarks are
presented. Finally, numerical simulations of hypotheses presented in this work and comparison
between them are presented.

\section{Mass transfer definitions}
By the definition the mass transfer coefficient from the surface with the imposed constant
concentration $\cbubble$ is the following:
\beq
k_L=\frac{\dot{m}}{P \Delta C},
\feq
where $\dot{m}$ is the mass rate $\Bigl[\frac{kg}{s}\Bigr]$, $P$ is the area of the surface
$\Bigl[m^2\Bigr]$, $\Delta C$ is the concentration difference between the surface and the medium
$\Bigl[\frac{kg}{m^3}\Bigr]$. Therefore, $k_L$ has a dimension of velocity
$\Bigl[\frac{m}{s}\Bigr]$. Usually, the medium concentration is taken at the infinite distance
from the bubble. However, in the case bubble, the medium concentration can be the average
concentration in the domain, the flux averaged concentration at the outlet. Thus, one needs to
establish a thorough definition of the volumetric mass transfer coefficient in the case of complex
geometries and hydrodynamic patterns.

>>>>

\newcommand{\beq}{\begin{equation}}
\newcommand{\feq}{\end{equation}}
\newcommand{\beqstar}{\begin{equation*}}
\newcommand{\feqstar}{\end{equation*}}                                     
\newcommand{\beqal}{\begin{equation}\begin{aligned}}
\newcommand{\feqal}{\end{aligned}\end{equation}}
\newcommand{\beqalstar}{\begin{equation*}\begin{aligned}}
\newcommand{\feqalstar}{\end{aligned}\end{equation*}}
\newcommand{\vol}{k_L\,a}
\newcommand{\lbubble}{L_{\mathrm{bubble}}}
\newcommand{\lunit}{L_{\mathrm{unit}}}
\newcommand{\lslug}{L_{\mathrm{slug}}}
\newcommand{\ububble}{U_{\mathrm{bubble}}}
\newcommand{\uliq}{U_{\mathrm{liq}}}
\newcommand{\ugas}{U_{\mathrm{gas}}}
\newcommand{\uoutlet}{U_{\mathrm{outlet}}}
\newcommand{\cbubble}{C_{\mathrm{bubble}}}
\newcommand{\cinlet}{C_{\mathrm{inlet}}}
\newcommand{\coutlet}{C_{\mathrm{outlet}}}

@@

\begin{document}
\maketitle
\begin{abstract}
This work presents a thorough procedure of how to determine the volumetric mass transfer
coefficient
in the context of lattice Boltzmann simulations. The novelty of this work lies in the establishing
a thorough procedure of how to measure the volumetric mass transfer coefficient in case of mass
sources to be comparable of the representative unit cell, which accounts for the inhomogenous
situation. By examining $5$ representative cases taken from microfluidics with the capillary number
from $Ca=0.1$ to $Ca=1$ the following relationship was established {\color{red} Give main findings
here}. The work is of use to people performing mass transfer numerical simulations.
\end{abstract}

\section{Introduction}
The volumetric mass transfer coefficient is a very important characteristic for microchannels 
design \cite{kreutzer-overview}. It allows to
properly manufacture a microchannel with necessary properties to ensure that chemical
reactions are performed in the best manner. Many chemical microchannel configurations
\cite{kreutzer-pressure-drop}  The flow for microchannels we focus on is the bubble train
motion, where bubbles organized in a train are traveling through the liquid media. This
configuration is interesting due to several reasons: hydrodynamic velocity patterns are completely
different for different capillary numbers, the bubble occupies the significant part of the unit cell
and the flow can be assumed periodic.

For the bubble train the volumetric mass transfer coefficient correlations are represented as a
function of the diffusion coefficient, slug and bubble lengths, and bubble velocity. For example,
\citet{yue-mass} established an experimental correlation for a bubble train as: 
\begin{equation}
\vol =\frac{2}{d_h} \Bigl(\frac{D
\ububble}{\lbubble+\lslug}\Bigr)^{0.5}
\Bigl(\frac{\lbubble}{\lbubble+\lslug}\Bigr)^{0.3},
\end{equation}
where $\vol$ is the volumetric mass transfer coefficient, $d_h$ is the hydrodynamic capillary
diameter, $\lbubble$ is the bubble length, $\lslug$ is the slug distance (between bubbles),
$\ububble$ is the bubble velocity. This correlation is within $10\%$ accuracy for experimental
results for the following range of
parameters: bubble velocity $0.4\,\mathrm{m/s}<\ububble<2\,\mathrm{m/s}$, bubble length
$1.4<\frac{\lbubble}{d_h}<6.3$, slug length $1<\frac{\lslug}{d_h}<3.2$,
hydrodynamic diameter $d_h = 400 \mathrm{\mu m}$.

Other correlations can be found in works of \citet{kreutzer-overview}
and \citet{bercic-mass}. Among a variety of experimental works extensively covered
\cite{yue-mass}, the numerical validations to the best authors' knowledge are sparse. One of the
reasons is that the experimental correlations are established for continous system, i.e. one does
not distinguish separate bubbles, but looks to the system as a whole measuring mass concentration
in different places. Thus, numerical simulations require a significant computational
resources to be able to simulate enough number of the unit cells to mimic the continuous picture.
There are some works as how to avoid large scale numerical simulations by performing numerical
simulations for one unit cell. For example, \citet{vanbaten-circular} perform numerical
simulations for one unit cell using periodic boundary conditions and measure the volumetric mass
transfer coefficient as:
\begin{equation}
\label{main:simulation:equation}
k_L a=\frac{\mathrm{Flux}}{\cbubble-\coutlet} \frac{\mathrm{bubble\ surface\ area}}{\mathrm{unit\
cell\ volume}},
\end{equation}
where $\coutlet=\int{C \uoutlet \mathrm{d}A}/\int{\uoutlet\mathrm{d}A}$. 
The concentration flux is calculated as the difference between overall
average concentration in the whole domain ($C_{overall}=\int_{V} C \mathrm{d}V /V$)
at time
$t_1$ and at time $t_2$ divided on the time difference $t_2-t_1$. The agreement between numerical
simulations and correlations of \citet{bercic-mass} was good. However, the numerical model only
accounts for circular capillaries and short contact times when the
bubble film is not saturated with tracer. As well, authors took the
approximation of bubble shapes for small capillary numbers $Ca<0.01$ and then derived the
hydrodynamics fields by imposing a free-slip condition at the bubble surface. {\color{red} Give
another numerical work by Kreutzer.}

This work aims at the establishing numerical procedures as to how properly obtain the volumetric
mass transfer coefficient in the range of capillary number $0.1\leq Ca\leq 1.0$, where the
hydrodynamic velocity pattern changes. Different boundary conditions, i.e. the periodic boundary
conditions, open boundaries conditions will be examined to determine the volumetric mass transfer
coefficient. As well, large scale simulations are performed, which simulate from 4 to 10 unit
cells.

The comparison of volumetric mass transfer coefficient
correlations with experimental correlations and establishing the Sherwood number as a function of
other non-dimensional parameters is the goal of the future work. However, as an example we will
show that one of the most recent correlations for the volumetric mass transfer coefficient by
\citet{yue-mass} resembles current simulations.

We chose to simulate mass transfer of the bubble train due to different reasons:
\begin{description}
\item[I] The gas holdup, i.e. the ratio of volume occupied by bubble to unit cell volume, is
non-negligible. Therefore, one cannot assume that the mass source is a point source which accounts
to the homogenous advection-diffusion equation with the mass source.  Therefore, there is a logical
mismatch: on one hand one needs to account the complicated geometry of the bubble and account for
all velocity pattern details, on the other hand one needs to mimic the continous picture which is
seen in experiments. One of the goals of this work is to establish certain criteria as how to
simulate such problems numerically with the help of the lattice Boltzmann method.
\item[II] For small capillary numbers as $Ca<0.7$ \cite{giavedoni-numerical} there is a vortex in
the reference frame moving with
the bubble, which  influences and improves the corresponding mass transfer.  For larger capillary
numbers there is no vortex and stremalines go around the bubble {\color{red} Give two pictures
with corresponding streamlines}. Thus, the problem is interesting because of different
hydrodynamic patterns depending on the capillary number which governs a number of experimental
parameters as the slug length, the film thickness and the existence of the vortex in front of the
bubble. There are some available
correlations in literature \cite{bercic-mass,kreutzer-overview}. One of the particular goal of
this work is to study mass transfer dependence on the completely different velocity patterns
depending on the capillary number $Ca$. 
\end{description}

The numerical approach we take is the lattice Boltzmann method, a relatively new CFD
competitor emerged during last $20$ years \cite{frisch,mcnamara,HJ,HSB}. During years the
method was applied to simulate not only hydrodynamic problems \cite{yu}, but as well multiphase
flows \cite{Shan-chen:extended,swift,gunstensen}, heat transfer
\cite{yuan-thermal,zhang-thermal}, ferrofluids \cite{dellar-ferro,kuzmin-aniso}.

The mass transfer problems in the lattice Boltzmann framework were mainly addressed in series of
works of Dr. Ginzburg and co-authors
\cite{ginzburg-main,ginzburg-boundary-conditions,ginzburg-saturated-flow}. However, all these works
are of general nature to simulate the advection-diffusion equation via the lattice Boltzmann
framework. In comparison, this work aims to establish the procedure of how to obtain the
volumetric mass transfer coefficients for bubble train flow. One should mention the work of
\citet{inamuro-scalar-boundary} simulates heat and mass transfers in the porous media and the work
of \citet{jos-mass} simulates lateral mixing in cross-channel flow. While two last works are focused
at the particular mass transfer problems, both problems are of homogenuous nature and do not
guide as how to simulate the non-homogenous problems.

The overall goal of this paper is to address the questions arised and to establish the numerical
procedure suitable to measure the volumetric mass transfer coefficient in the context of the
lattice Boltzmann method. We will utilize the results of numerical simulations of bubble flow
obtained
before \cite{kuzmin-binary2d,kuzmin-binary3d}.

The paper is organized as follows. The mass transfer coefficients and volumetric mass transfer
coefficients definitions are covered. The limiting cases from the continuos representation and their
application to present simulations are given. The mass transfer simulations benchmarks are
presented. Finally, numerical simulations of different boundary conditions and large scale
simulations and comparison between them are presented.

\section{Mass transfer definitions}
By the definition the mass transfer coefficient from the surface with the imposed constant
concentration $\cbubble$ is the following:
\beq
k_L=\frac{\dot{m}}{P \Delta C},
\feq
where $\dot{m}$ is the mass rate $\Bigl[\frac{kg}{s}\Bigr]$, $P$ is the area of the surface
$\Bigl[m^2\Bigr]$, $\Delta C$ is the concentration difference between the surface and the medium
$\Bigl[\frac{kg}{m^3}\Bigr]$. Therefore, $k_L$ has a dimension of velocity
$\Bigl[\frac{m}{s}\Bigr]$. Usually, the medium concentration is taken at the infinite distance
from the bubble. However, in the case of complicated geometries and inhomogenuous concentrations, 
the medium concentration can be the average concentration in the domain, the flux averaged
concentration at the outlet. Thus, one needs to establish a thorough definition of the volumetric
mass transfer coefficient in the case of complex geometries and hydrodynamic velocity patterns.
